Data Specialization for a Staged Lambda Calculus

### 2014-11-06 paper/splitting.tex
pass over splitting
fixed figures
some work with nico

Given a term in the source language \lang\ where terms of
stage~\bbone\ and stage-\bbtwo\ may be interleaved, \emph{(stage)
  splitting} separates the stage-\bbone\ and stage-\bbtwo\
subcomputations into two separate terms, each performing the
subcomputations corresponding to each stage.  More precisely, given a
\lang\ term~$e$, splitting yields two terms $e_1$ and $e_2$ in a
unstaged target language \langmono\ such that~$e_1$ performs all the
stage~\bbone\ subcomputations in~$e$ and $e_2$~performs all of the
stage~\bbtwo\ subcomputations.

\ur{Technically, target language is not lambda but lambda plus pairs.
  Please use some other symbol than lambda, lambda-p lambda+ or
  something like that, or simply:} \textsf{\bf \textsc{Mono}}.

Splitting preserves correctness in the sense that evaluation of~$e$ in
the staged language produces the same result as evaluating~$e_1$ and
then~$e_2$ with the result from~$e_1$.  In this section, we present a
splitting algorithm for the source language \lang\, specifying
precisely the algorithm and the correctness condition.  We start by
defining a masking relation that helps us to isolate stage-1 and
stage-2 components of a partial value and then move on to the
specification of the splitting algorithms.

@@

{\em Partial-value masking} or simply {\em masking} are a functions
for each stage that project from the domain of a partial value and
residual table to the a term in the target language \langmono.

\ur{I am attempting to define this even though it is not clear from
  the text what it is.  The definition should specify precisely the
  domain and the range of the functions.  Does the domain include the
  table?  The figure does not.  }

\ref{fig:valMask} defines the functions, $\masko{\cdot}$ and
$\maskt{\cdot}$. Since \texttt{next}'s only contain references to the
residual table (computations that occur in stage~\bbtwo),
$\masko{\xi;v}$ is to replaces all~$\next$s in~$v$ with~\texttt{()}.
The result is a \langmono\ value with the same structure as~$v$, but
containing only its stage~\bbone\ information. Masking is more complex
when~$v$ is a function since the function's body itself may contain
stage~\bbtwo\ subterms.  As a result, masking of functions depends on
the general behavior of stage~\bbone\ splitting, described shortly.

While $\masko{\cdot}$ produces a value in \langmono, $\maskt{\xi;v}$
produces a term in \langmono\ that corresponds to the stage~\bbtwo\
computation represented by~$[\xi;v]$.  Stage~\bbone\ information
in~$v$ is discarded by replacing base constants with~\texttt{()},
replacing $\next~\hat{y}$ with~$\hat{y}$, and eliminating all
injection tags. (The sum type is at stage \bbone, so the tag is not
relevant at stage \bbtwo.)  Last, $\maskt{\cdot}$ reifies~$\xi$ around
the modified~$v$ to produce a term in \langmono.  As with
$\masko{\cdot}$, function masking by $\maskt{\cdot}$ depends on
general term splitting.

\subsection{Splitting Algorithm}

\paragraph{Splitting of stage-\bbone\ terms.}

@@

OUTLINE:

\begin{itemize}
\item 
The structure of the splitting judgment.

\item
The equivalence condition.

\item
Walk through the cases.
\end{itemize}

\paragraph{Splitting stage-\bbtwo\ terms.}

\subsection{Splitting Stage \bbone\ Terms}

@@

The goal of stage 2 splitting was to generate a precomputation and resumer such that the resumer and the \lang\ residual reduce to the same final value. However, stage~\bbone\ terms do not reduce to final values, rather they reduce to residual tables and partial values.  For example, consider the stage 1 term: 
\begin{lstlisting}
1`(1+2, next{`2`hold{`1`3+4`2`} + 5`1`})`
\end{lstlisting}
which evaluates (via $\mathbin{\redonesym}$), to the following residual table and a multi-stage tuple:

\begin{lstlisting}
2`[yhat|->7+5]` 1`(3,next{`2`yhat`1`})`
\end{lstlisting}

@@

Note that this result simultaneously represents information at {\em both} stages: 
a stage~\bbone\ value 3, and the stage~\bbtwo\ computation $\letin \yhat {\mathtt{7+5}} \yhat$. 
Therefore, to make the results of splitting match the the partial value above, we choose to split \bbone\ terms in \lang\ into a resumer that reduces to the same value as the stage~\bbtwo\ computation, and an $e1$ that not only performs the precomputation expected by the resumer, but also computes values that match the stage~\bbone\ aspects of the partial value.

@@

\subsubsection{Splitting Mechanics}

@@

>>>>

Given a term in the source language \lang\ where terms of
stage~\bbone\ and stage~\bbtwo\ may be interleaved, \emph{(stage)
  splitting} separates the stage~\bbone\ and stage~\bbtwo\
subcomputations into two separate terms, each performing the
subcomputations corresponding to each stage.  More precisely, given a
\lang\ term~$e$, splitting yields two terms $e_1$ and $e_2$ in a
unstaged target language \langmono\ such that~$e_1$ performs all the
stage~\bbone\ subcomputations in~$e$ and $e_2$~performs all of the
stage~\bbtwo\ subcomputations.

Splitting preserves correctness in the sense that evaluation of~$e$ in
the staged language produces the same result as evaluating~$e_1$ and
then~$e_2$ with the result from~$e_1$.  In this section, we present a
splitting algorithm for the source language \lang\, specifying
precisely the algorithm and the correctness condition.  We start by
defining a masking relation that helps us to isolate stage~\bbone\ and
stage~\bbtwo\ components of a partial value and then move on to the
specification of the splitting algorithms.

@@

Intuitively, we require that the terms obtained by splitting a term $e$, when
evaluated, agree with $e$ when evaluated according to our dynamic semantics. In
the case of stage~\bbone\ terms, our dynamic semantics produce partial values
containing information at \emph{both} stages, which goes against the idea that
splitting separates the subcomputations belonging to each stage.

For example, consider the stage 1 term
\begin{lstlisting}
1`(1+2, next{`2`hold{`1`3+4`2`} + 5`1`})`
\end{lstlisting}
which evaluates via $\redonesym$ to a residual table and multi-stage tuple
\begin{lstlisting}
2`[yhat|->7+5]` 1`(3,next{`2`yhat`1`})`
\end{lstlisting}
This partial value contains both a stage~\bbone\ value 3, and a stage~\bbtwo\
computation $\letin \yhat {\mathtt{7+5}} \yhat$.

To clearly delineate these parts of the partial value, we define 
\emph{partial-value masking}, or simply {\em masking}, functions at each stage,
which compute from a residual table and partial value the content (expressed as
a term in \langmono) belonging to each stage. \ref{fig:valMask} defines these
masking functions, written $\masko{{-}}$ and $\maskt{{-};{-}}$.

The goal of $\masko{v}$ is to produce a term in \langmono\ containing exactly
the stage~\bbone\ content of $v$. Since \next\ terms only contain stage~\bbtwo\
content, we replace all $\next$ subterms of $v$ with \texttt{()}.  For lambdas,
since these represent potentially-multistage computations, we split their bodies
as general stage~\bbone\ terms (as described in \ref{sec:split-one}), and return
the stage~\bbone\ component. $\masko{v}$ is always a value in \langmono, and has
the same structure as~$v$.

The goal of $\maskt{\xi;v}$ is to produce a term in \langmono\ containing
exactly the stage~\bbtwo\ content of $[\xi;v]$. Unlike in stage \bbone\ masking,
here we need access to the residual table $\xi$, because all stage~\bbtwo\
content in $v$ has already been lifted out into $\xi$. To discard all
stage~\bbone\ information in $v$, we replace base constants with~\texttt{()},
replace $\next~\hat{y}$ with~$\hat{y}$, and replace all $n$-ary term
constructors with an $n$-tuple of their subterms, themselves masked. As before,
we mask functions by splitting their bodies as general stage~\bbone\ terms, and
use the stage~\bbtwo\ component.

After performing this transformation on $v$, we reify $\xi$ around the result,
producing a term in \langmono.

\subsection{Splitting Stage \bbone\ Terms}
\label{sec:split-one}

@@

FOLLOW OUTLINE:

\begin{itemize}
\item The structure of the splitting judgment.
\item The equivalence condition.
\item Walk through the cases.
\end{itemize}

@@

@@

@@

@@

The goal of stage 2 splitting was to generate a precomputation and resumer such
that the resumer and the \lang\ residual reduce to the same final value.
However, stage~\bbone\ terms do not reduce to final values, rather they reduce
to residual tables and partial values.  
Therefore, to make the results of splitting match the
the partial value above, we choose to split \bbone\ terms in \lang\ into a
resumer that reduces to the same value as the stage~\bbtwo\ computation, and an
$e_1$ that not only performs the precomputation expected by the resumer, but
also computes values that match the stage~\bbone\ aspects of the partial value.

\subsection{Splitting Stage \bbtwo\ Terms}
